A Model to Predict Survival in Patients with End-Stage Liver Disease

### 2021-08-03 analysis/article.Rmd
feat: add flowchart and discussion notes

---
editor_options:
  chunk_output_type: console
title: "A Model to Predict Survival in Patients with End-Stage Liver Disease"
subtitle: >
    Comparision against MELD, MELD-Na, MELD-Plus7.
abstract: >
  **Background**:
      TODO

@@

author:
    - Sebastian Gibb:
        email: mail@sebastiangibb.de
        institute: [ILM, KAINS]
        correspondence: true
    - Thorsten Kaiser:
        email: thorsten.kaiser@medizin.uni-leipzig.de
        institute: ILM
        correspondence: false
institute:
    - ILM: Institute of Laboratory Medicine, Clinical Chemistry and Molecular Diagnostics, University Hospital Leipzig, Paul-List-Str. 13-15, D-04103 Leipzig, Germany.
    - KAINS: Anesthesiology and Intensive Care Medicine, University Hospital Greifswald, Ferdinand-Sauerbruch-Straße, D-17475 Greifswald, Germany.
date: "`r Sys.Date()`"
bibliography:
  - "`r rprojroot::find_rstudio_root_file('analysis', 'bibliography', 'bibliography.bib')`"
  - "`r rprojroot::find_rstudio_root_file('analysis', 'bibliography', 'rpackages', 'article.bib')`"
output:
  workflowr::wflow_html:
    pandoc_args:
      - --filter=pandoc-crossref
        # pandoc_args doesn't support r evaluation
      - "--lua-filter=pandoc/lua-filters/scholarly-metadata.lua"
      - "--lua-filter=pandoc/lua-filters/author-info-blocks.lua"
---

```{r, include=FALSE}
knitr::write_bib(
    c("targets", "workflowr"),
    file = rprojroot::find_rstudio_root_file(
        "analysis", "bibliography", "rpackages", "article.bib"
    )
)
```

@@

@@

>>>>

---
# R markdown template taken from
# Dan Holmes <dtholmes@mail.ubc.ca>
# https://labrtorian.com/2019/08/26/rmarkdown-template-that-manages-academic-affiliations/
title: "A Model to Predict Survival in Patients with End-Stage Liver Disease"
subtitle: >
    Comparision against MELD, MELD-Na, MELD-Plus7.
abstract: >
  **Background**:
      TODO

@@

author:
    - Sebastian Gibb:
        email: mail@sebastiangibb.de
        institute: [ILM, KAINS]
        correspondence: true
    - Thorsten Kaiser:
        email: thorsten.kaiser@medizin.uni-leipzig.de
        institute: ILM
        correspondence: false
institute:
    - ILM: Institute of Laboratory Medicine, Clinical Chemistry and Molecular Diagnostics, University Hospital Leipzig, Paul-List-Str. 13-15, D-04103 Leipzig, Germany.
    - KAINS: Anesthesiology and Intensive Care Medicine, University Hospital Greifswald, Ferdinand-Sauerbruch-Straße, D-17475 Greifswald, Germany.
date: "`r Sys.Date()`"
bibliography:
  - "`r rprojroot::find_rstudio_root_file('analysis', 'bibliography', 'bibliography.bib')`"
  - "`r rprojroot::find_rstudio_root_file('analysis', 'bibliography', 'rpackages', 'article.bib')`"
output:
  bookdown::html_document2:
    pandoc_args: # pandoc_args doesn't support r evaluation
      - --lua-filter=pandoc/lua-filters/scholarly-metadata.lua
      - --lua-filter=pandoc/lua-filters/author-info-blocks.lua
---

```{r knitr_setup, include=FALSE}
knitr::write_bib(
    c("targets", "workflowr"),
    file = rprojroot::find_rstudio_root_file(
        "analysis", "bibliography", "rpackages", "article.bib"
    )
)
```

```{r article_setup, include=FALSE}
library("targets")

tar_load(raw_data)
tar_load(bmrk_results)
```

@@

## Study population

```{r excluded, include = FALSE}
n_excluded <- attr(raw_data, "excluded")
n_analysed <- nrow(raw_data)
n_total <- sum(n_analysed, n_excluded)

cn_labs <- grep("_[CESQ]$", colnames(raw_data), value = TRUE)
```

In a retrospective cohort study we followed `r n_total` consecutive
patients, who were recruited during the evaluation process for
liver transplantation at the University Hospital of Leipzig from
`r paste(format(attr(raw_data, "period"), "%B %Y"), collapse = " to ")`.
For each patient we recorded `r ncol(raw_data)` variables.
Among them are age, sex, etiology of liver disease, complications as listed in
Table \@ref(tab:tableone) and `r length(cn_labs)` laboratory measurements.

```{r flowchart, results = "asis", echo = FALSE, fig.width = 5, fig.height = 5, fig.cap = "Flowchart Inclusion/Exclusion."}
graph <- "
digraph flowchart {
    graph [splines = false];

    // invisible nodes
    node [shape = point, width = 0, height = 0];
    iAR [style = 'invis'];

    // visible nodes
    node [shape = box];
    A [label = '@@1'];
    E [label = '@@3'];
    R [label = '@@2'];

    // edge statements
    A -> iAR [arrowhead = none];
    {rank = same; iAR -> E [minlen = 5]};
    iAR -> R;
}

[1]: paste0('Total Records\\n(n = ', n_total, ')')
[2]: paste0('Analysed Records\\n(n = ', n_analysed, ')')
[3]: str_excluded
"
str_excluded <- paste0(
    'Excluded (n = ', sum(n_excluded), ')\\\\l',
    paste0(
        '- ',
        paste(
            c(
                'Younger than 18 years:',
                'Liver Transplantation before visit:',
                'Follow-up missing/invalid:'
            ),
            n_excluded
        ),
        collapse = '\\\\l'
    ),
    '\\\\l'
)

DiagrammeR::grViz(graph)
```

We excluded `r sum(n_excluded)` patients from our analysis
(Figure \@ref(fig:flowchart)).
`r if (n_excluded["Children"] == 1) "One was" else paste(n_excluded["Children"], "were")` younger than 18 years.
`r n_excluded["LiverTransplantation"]` had a liver transplantation before and
in `r n_excluded["LostFollowUp"]` cases the follow up data were missing or
invalid.

The Ethics Committee at the Leipzig University Faculty of
Medicine approved the retrospective usage of the data for our study
(reference number: 039/14ff).

## MELD scores

MELD and MELD-Na were calculated as described in
@optn_policies2021 using the following formulas:
$MELD = 10 * (0.957 \ln(creatinine [mg/dl]) + 0.378 \ln(bilirubin [mg/dl]) + 1.120 * \ln(INR) + 0.643)$,
Creatinine, bilirubin and INR values lower than 1.0 mg/dl were set to 1.0 mg/dl.
The maximum allowed creatinine was 4.0 mg/dl. If patients received dialysis,
creatinine was set to 4.0 mg/dl.
For MELD-Na MELD was calculated as above and recalculated if greater than 11 using:
$MELD\text{-}Na = MELD_i + 1.32 * (137 - Na [mmol/l]) - (0.033 * MELD_i * (137 - Na [mmol/l]))$

Serum sodium values lower than 125 mmol/l or higher than 137 mmol/l were set
to 125 mmol/l and 137 mmol/l, respectively.

The MELD-Plus7 risk score was calculated as described in @kartoun2017:
$$
\begin{aligned}
L = 8.53499496 &+ \\
    2.59679650 &* \log_{10}(1 + creatinine [mg/dl]) + \\
    2.06503238 &* \log_{10}(1 + bilirubin [mg/dl]) + \\
    2.99724802 &* \log_{10}(1 + INR) - \\
    6.47834101 &* \log_{10}(1 + sodium [mmol/l]) - \\
    6.34990436 &* \log_{10}(1 + albumin [g/l]) + \\
    1.92811726 &* \log_{10}(1 + wbc [th/cumm]) + \\
    0.04070442 &* age [years]
\end{aligned}
$$

$$MELD\text{-}Plus = \frac{\exp(L)}{1 + \exp(L)}$$

The MELD-Plus9 score was ignored because the length of stay information was not
available for our cohort and the model itself wasn't superior to MELD-Plus7 in
its original publication [@kartoun2017].

## Machine learning

```{r ml, include = FALSE}
n_lrns <- nrow(bmrk_results$learners)
#n_cv <- c(inner.folds
#    outer = unlist(
#        bmrk_results$resamplings$resampling[[1]]$param_set$get_values()
#    )
#)
```

@@

In this retrospective analysis of `r n_analysed` consecutive patients,
who were recruited during the evaluation process for liver transplantation at
the University Hospital of Leipzig we compared
`r n_lrns` different statistical and machine learning algorithms.

...

- higher mortality than predicted, compare @hernaez2020
- adding sodium doesn't add much/any discrimination information
    - as already seen in @kartoun2017
- @kartoun2017 uses > 60 variables mostly clinical stuff, just few laboratory values, none of the clinical stuff remains
- general low sample size thus simpler models/fewer predictors are preferred.
- comparison of different predictors of MELD-Plus7 (+MELD-Plus9) vs our features
    - 90 day mortality *post* discharge
    - poorer performance in own validation set (lower and more uniform disease severity, also lower MELD)
- machine learning algorithms worse than statistical learning (lower c-index, higher variance/instability)
    - additive effects (favour statistical models)
    - low sample size, for ML more than 200 events per predictor variable necessary [@ploeg2014]
- national register required

## Limitations

- comparable mortality?
- retrospective analysis
- single center
- unknown cause of death (death with vs of end-stage liver disease)
- bias caused by LTx censorship?
